---
title: "Lecture 17"
subtitle: "Statistical inference"
---


# Function to create a simulated dataset 

```{r}
library(tidyverse)
library(ggplot2)
library(broom)

# Function to create dataset with N observations
data_sample <- function(N = 100, alpha = 0, beta = 1, sd_e = 4) {
  data <- tibble(
    x = runif(N, 0, 1),
    y = alpha + beta*x + rnorm(N, mean=0, sd=sd_e) )
  return(data)
}
```

---

# Create simulated dataset with N=100 observations

```{r}
set.seed(1) # Set a seed because the error term is random
df100 <- data_sample(N=100)

nrow(df100)
head(df100)

```

---
# Try it: estimate beta using OLS

```{r, eval=F}
# Estimate y = alpha + beta*x + e
lm1 <- 
  
# What is the standard error of beta? p-val and 95% confidence interval?
# Is the estimate statistically significant at 95% confidence level?
tidy(lm1, conf.int = T, conf.level = 0.95)
```


---
# Approximate the 95% confidence interval

```{r, eval=F}

beta <- lm1$coefficients["x"]
stderr <- sqrt(diag(vcov(lm1)))["x"]

# 95% confidence interval is approximately beta +/- 2 stderrs
beta - 2*stderr
beta + 2*stderr

```

---

# What happens to estimates when sample size increases?

```{r}
set.seed(1)
df1000 <- data_sample(N=1000)

# Did the estimate of beta change?
# What happened to std error, p-val, and confidence interval?
lm2 <- lm(y ~ x, data = df1000)
tidy(lm2, conf.int = T, conf.level = 0.95)
```

---

# Write a function to automate the estimation of beta

```{r}

# Create a random dataset, and return the estimate of beta
estimate_beta <- function(N = 100) {
  df <- data_sample(N)
  lm <- lm(y ~ x, data = df)
  return(lm$coefficient["x"])
}

set.seed(1)
estimate_beta()

```

---

# Create 1000 datasets and estimate the betas

```{r}
N <- 100
ndatasets <- 1000

# Recall: lapply() is like a for loop
set.seed(1)
betas <- lapply(1:ndatasets, function(i) estimate_beta(N)) %>% 
  bind_rows()

nrow(betas)

```

---

# Create 1000 datasets and estimate the betas


```{r}

ggplot(betas, aes(x=x)) + 
  theme_bw() + 
  geom_histogram(color="black", 
                 fill="skyblue")
```


---

# What happens if we increase the sample size?


```{r}
set.seed(1)
N <- 1000 #<<
betas$x2 <-lapply(1:ndatasets,
                function(i) estimate_beta(N)) %>% 
       bind_rows()

ggplot(betas) + 
  theme_bw() + 
  geom_histogram(aes(x=x), alpha=0.2,
                 color="black", 
                 fill="skyblue") + 
  geom_histogram(aes(x=x2$x), alpha=0.2,
                 color="black", 
                 fill="red")
```

