---
title: "Lecture 24"
subtitle: "Regression discontinuity estimation"
---

# Load the data

```{r, message=F}
library(tidyverse)
library(broom)
df <- read_csv("exam.csv")

head(df)

```

---

# Try it: scatterplot of the data

```{r , eval=F}
# Scatterplot of Score (x) vs Income (y)
# Use geom_point() for the scatter plot
# Use geom_vline() to add vertical line at 
#    xintercept=50
library(ggplot2)
ggplot(  , aes( )) +
 

```


---

# Try it: local (non-parametric) RD

- Calculate Difference in Means using just the `50-54` and `45-49` data points


```{r, eval=F}

# Income (Score = 50) - Income(Score = 45)
diff_means <- 

```


---

# Try it: global (parametric) RD with no curve

- Calculate Difference in Means using **all data points**


```{r, eval=F}
# Create a Treat variable equal to 1 if Score>=50, 0 otherwise
df$Treat <- 

# Mean income (Treat=1) - mean income (Treat=0)


```


---

# Global (parametric) RD with no curve

- Can also be calculated using linear regression

$$Income = \alpha + \rho Treat + \epsilon$$

```{r c3, eval=F}
# regress Income on treat
RD <- lm(Income ~ Treat, data=df)
tidy(RD)
```


---

# Estimate an RD with a linear curve

$$Income = \alpha + \rho Treat + \gamma Score + \epsilon$$

- Is the estimate statistically significant?

--

```{r, eval=F}

# regress Income on Treat and Score
RD_lin <- lm(Income ~ Treat + Score, data=df)
tidy(RD_lin)

```



---

# Let the linear slope differ on either side of cutoff

$$Y = \alpha + \rho T + \gamma_1 X + \gamma_2 T\times (X - Cut) + \epsilon$$

```{r c4, eval=F}
df$TreatScore50 <- df$Treat*(df$Score-50)
RD_lin_differ <- lm(Income ~ Treat + Score + TreatScore50, data=df)
tidy(RD_lin_differ)
```


---

# Use a bandwidth of 25

```{r c5, eval=F}
RD_lin_differ_bw25 <- lm(Income ~ Treat + Score + TreatScore50, 
                         data=subset(df,abs(Score-50) <= 25)) #<<
tidy(RD_lin_differ_bw25)
```


---

# Use a quadratic curve

```{r c6, eval=F}
# Does this affect the estimate much?
RD_quad <- lm(Income ~ Treat + poly(Score,2), data=df)
tidy(RD_quad)
```


---

# Compare estimates


```{r, eval=F}
print(paste("Local RD estimate:", round(diff_means,2)))
cat("\n")

estimates <- list(RD, RD_lin, RD_lin_differ, RD_lin_differ_bw25, RD_quad)
sapply(estimates, function(x) subset(tidy(x, conf.int=T), term=="TreatTRUE"))
```



