---
title: "Lab 15 - Neural networks"
author: Ives He
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: simplex
    number_sections: false
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
knitr::knit_meta(clean=T)
```

## 0 load the packages 
```{r}
library(neuralnet)
library(caret)

```

## 1 create a data frame
```{r}
# load the data 
df <- read.csv('/Users/iveshe/Library/Mobile Documents/com~apple~CloudDocs/Term 1/FIN 550/Lab/lab-15/EastWestAirlinesNN.csv')
# first six rows 
head(df)
# column names
names(df)
```

## 2 create two outcome dummies to represent output nodes  
```{r}
# create a variable named purchase
df$purchase <- df$Phone_sale == 1
# create a variable named not_purchase
df$not_purchase <- df$Phone_sale == 0

head(df[, c("Phone_sale", "purchase", "not_purchase")])
```

## 3 remove unnecessary variables
```{r}
# remove ID column and update df 
df <- df[, -1]
# column names 
names(df)
```

## 4 remove rows with missing values
```{r}
df <- na.omit(df)
dim(df)
```

## 5 data partition 
```{r}
# set the seed 
set.seed(10)
# row indexes of the training set
train.index <- sample(c(1:dim(df)[1]), 0.6 * dim(df)[1])
# training set
train.df <- df[train.index, ]
# test set
test.df <- df[-train.index, ]

dim(train.df)
dim(test.df)
```

## 6 normalize predictors to a scale of [0, 1] 
```{r}
# estimate the transformation 
norm.values <- preProcess(train.df[, 1:14], method = "range")
# normalize the numerical predictors in the training set 
train.norm.df <- predict(norm.values, train.df[, 1:14])
train.norm.df$purchase <- train.df$purchase
train.norm.df$not_purchase <- train.df$not_purchase
# normalize the numerical predictors in the test set 
test.norm.df <- predict(norm.values, test.df[, 1:14])
test.norm.df$purchase <- test.df$purchase
test.norm.df$not_purchase <- test.df$not_purchase

head(train.norm.df)
head(test.norm.df)
```

## 7 fit a neural network on the training set 
```{r}
# set the random seed 
set.seed(10)
# fit a neural network that has a single hidden layer with 5 nodes (should take ~ 1 minute)
nn <- neuralnet(purchase + not_purchase ~ Topflight + Balance + Qual_miles + cc1_miles. +
                cc2_miles. + cc3_miles. + Bonus_miles + Bonus_trans + Flight_miles_12mo +
                Flight_trans_12 + Online_12 + Email + Club_member + Any_cc_miles_12mo, 
                data = train.norm.df, hidden = 5, linear.output = FALSE)
# plot the network
plot(nn, rep = "best", 
     show.weights = TRUE,       
     information = FALSE,       
     intercept = FALSE,         
     cex = 0.8,                  
     circle.cex = 1.0,           
     col.hidden = 'blue',       
     col.entry = 'darkgreen',    
     col.out = 'darkred',        
     fontsize = 10,       
     arrow.cex = 1.2)  
# display weights
nn$weights
```

## 8 generate predicted probabilities for records in the test set  
```{r}
# make predictions using compute() 
nn.pred <- compute(nn, test.norm.df[, 1:14])
# predicted probabilities of making purchases
nn.pred.prob <- nn.pred$net.result[, 2] / (nn.pred$net.result[, 1] + nn.pred$net.result[, 2])
head(nn.pred.prob)
```

## 9 generate predicted classes for records in the test set 
```{r}
nn.pred.class <- ifelse(nn.pred.prob > 0.5, 1, 0)
head(nn.pred.class)
```

## 10 create a confusion matrix for records in the test set
```{r}
conf.matrix <- confusionMatrix(as.factor(nn.pred.class), as.factor(test.df$Phone_sale), positive = "1")
conf.matrix

```
